---
title: "Homework Assignment 1"
author: "Daniel Vasquez"
date: 'Assigned: Oct 24, 2020, Due Sun Nov 01, 2020 11:59PM'
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    
---

```{r}
library(dplyr)
```

## Reading my data

```{r, cache=TRUE}
nyc <- read.csv("nycflights.csv")

# Quick exploration of the data
str(nyc)
summary(nyc)
```

## Cleaning the data

```{r}
# Deleting the column X
nyc$X <- NULL

# Factorize columns that need to be a factor

nyc$origin <- as.factor(nyc$origin)
nyc$tailnum <- as.factor(nyc$tailnum)
nyc$month <- as.factor(nyc$month)
nyc$dest <- as.factor(nyc$dest)
nyc$carrier <- as.factor(nyc$carrier)
nyc$flight <- as.factor(nyc$flight)

# Re-check my data
str(nyc)
summary(nyc)
```

## Data Exploration

- How many airlines are there? (Hint: `levels` and `length` can be useful here)
- There are `r length(levels(factor(nyc$carrier)))` airlines.

```{r}
length(levels(factor(nyc$carrier)))
```

- How many flights were there by the airline with code `OO`? (Hint: `nrow` can be useful here along with logical indexing)
- `r nrow(nyc[factor(nyc$carrier) == "OO",])` flights were there by the airline with code `OO`.

```{r}
nrow(nyc[factor(nyc$carrier) == "OO",])
```

- How long is the shortest flight out of any NYC airport? (Hint: `min` can be useful, remember to handle `NA` values)
- The shortest flight out of any NYC airport is `r min(nyc$air_time, na.rm = TRUE)` minutes.

```{r}
min(nyc$air_time, na.rm = TRUE)
```

- How many flights were there by United Airlines (code: UA) on Jan 12th 2013?
- On January 12, 2013, there were `r nrow(nyc[nyc$carrier == "UA" & nyc$month == 1 & nyc$day == 12 & nyc$year ==2013, ])` flights by United Airlines.

```{r}
nrow(nyc[nyc$carrier == "UA" & nyc$month == 1 & nyc$day == 12 & nyc$year ==2013, ])
```

### Arrival Delay

- What was the average arrival delay for all airports and all airlines combined in Jan 2013?
- `r mean(nyc$arr_delay, na.rm = TRUE)`.

```{r}
mean(nyc$arr_delay, na.rm = TRUE)
```


- Whats was the median arrival delay for all airports and all airlines combined in Jan 2013?
- `r median(nyc$arr_delay, na.rm = TRUE)`.

```{r}
median(nyc$arr_delay, na.rm = TRUE)
```

What can you say about the distribution of arrival delays?
- The mean is larger number than the median, providing rationale that big delays were longer by a greater margin compared to both shorter delays & early arrivals.

### Airline Performance

- Calculate average arrival delays by airline.

```{r}
ave_arr_delays <- tapply(nyc$arr_delay, nyc$carrier, mean, na.rm = TRUE)
```

- Bar Plot of Average Arrival Delays for all the Airlines.

```{r}
barplot(sort(ave_arr_delays, decreasing = TRUE), cex.names = 0.6)
```

- Which airline has the highest average arrival delay? Which airline has the smallest average arrival delay? Are there airlines that actually have negative average delay?.

#Frontier Airlines had the highest average arrival delay.  Alaska Airlines had the shortest average arrival delay.  Native average delay means, I would imagine, that the flights make up more time in the air, on average, than they lose in departure delays on the ground, making them early more often than not. It's a relatively isolated city and they're a relatively small carrier with long flights. They have more opportunities to make up time in the air.

### Air Gain

Create a new column named airgain such that airgain = (departure delay - arrival delay) : this is the amount of delay a flight made up while in air.

```{r}
nyc$airgain <- nyc$dep_delay-nyc$arr_delay
```

- Explore airgain data - calculate suitable descriptive statistics and appropriate graphics to better understand this data. This part is open ended - you do what you feel works best for you.

- Do airlines actually gain any time when in air on average?
#Yes, airlines gain time when in the air, on average.

```{r}
meanairgain <- mean(nyc$airgain, na.rm = TRUE)
meanairgain

#If else statement
if (meanairgain > 0) { print("Airline gained time in the air")} else { print("No, negative airgain")}
```
  
  - Calculate average airgain for different airlines - which airlines do a better job, which do a worse job?
  - See explanation below chart.
  
```{r}
avgairgainbyairline <- tapply(nyc$airgain, nyc$carrier, mean, na.rm = TRUE)
avgairgainbyairline
```

```{r}
#Visualization to answer pt2
barplot(sort(avgairgainbyairline, decreasing = TRUE), cex.names = 0.6)
```
# This visualization shows which airlines are good at catching up when in air and which airlines actually lose time further after taking off. AS, HA, and VX do the best while MQ, FL, and F9 do the worst.

### Merging Data Frames

-Read the data.

```{r}
planes <- read.csv("planes.csv")

# Deleting the column X.
planes$X <- NULL

# Factorize columns that need to be a factor

planes$tailnum <- as.factor(planes$tailnum)
planes$type <- as.factor(planes$type)
planes$manufacturer <- as.factor(planes$manufacturer)
planes$model <- as.factor(planes$model)
planes$engine <- as.factor(planes$engine)
planes$engines <- as.factor(planes$engines)

# Re-check my data
str(planes)
summary(planes)

```
- Merge the flights data and the planes data.

```{r}
nycplanes <- merge(x = nyc, y=planes,
                   by = "tailnum", all = TRUE)
```

## Exploratory Analysis

- What is the 90th percentile of airgain for different manufacturers?
- See chart below for the list in descending order.

```{r}
nycplanes %>%
  group_by(manufacturer) %>%
  summarise(airgain90=quantile(airgain, 0.9, na.rm=TRUE)) %>%
  arrange(desc(airgain90))
```

 - What is the breakdown for engines in planes from LaGuardia Airport? How many planes are in each type of group?
 - Aircraft may have 1-4 engines, see chart below for number of planes in each group cutiepie!;)
 
```{r}
nycplanes %>%
  filter(origin == 'LGA') %>%
  group_by(engines) %>%
  summarize(number_of_planes=n())
```
 
- How many seats are in the second largest capacity aircraft? How many of these planes are in circulation?
- 400 seats and 29 aircrafts are in circulation.

```{r}
nycplanes %>%
  group_by(seats) %>%
  summarize(count = n()) %>%
  arrange(desc(seats)) %>%
  slice(2)
```

- What is the mean and median speed for American Airline flights with a destination at O'Hare International Airport?
- For speed, the mean is 108.8 and the median is 95.

```{r}
nycplanes %>%
  filter(carrier == 'AA',
         dest == 'ORD') %>%
  summarize(mean = mean(speed, na.rm = TRUE), median = median(speed, na.rm=TRUE))
```

- Should I fly in morning (4am - 12Noon), afternoon (12Noon - 4pm), evening (4pm - 8pm) or night (8pm - 4am) if I want to experience the least amount of arrival delay? [Use piping and 'mutate' command]
- Flying in the morning results in fewer delays.

```{r}
nycplanes %>%
  mutate(timeofday = 
           case_when(nycplanes$hour >= 4 & nycplanes$hour <= 12 ~'morning',
                     nycplanes$hour > 12 & nycplanes$hour <= 16 ~'afternoon',
                     nycplanes$hour > 16 & nycplanes$hour <= 20 ~'evening',
                     nycplanes$hour > 20 & nycplanes$hour <= 24 ~'night')) %>%
  group_by(timeofday) %>%
  summarize(mean_of_arrival_delay = mean(arr_delay, na.rm=TRUE))
```